Unfolding the genome's 3D architecture:\\ \Large{Insights in {\em P. falciparum} genome organization}

### 2017-07-28 doc/manuscript/paper.tex
more fixes

\begin{abstract}
\begin{itemize}

\item {\em P. falciparum} is the deadliest parasite causing human malaria,
responsible for more than 500,000 death in 2015. Our lack of understanding of
the parasite gene regulation of its complex life cycle is an impediment to the
development of novel therapies. Recent evidence points towards unusual means
of gene regulation, possible through changes in its genome architecture. A
better understanding of how DNA folds into the nucleus and how it influences
gene expression could provide new leads for developing new drugs.
\item In the meantime, the development of new means to probe samples for the
three-dimensional structure of DNA paves the way for in depth and systematic analysis of genome
architecture, offering new insights in many biological processes. 3C-like
methods can now assess physical interactions between pairs of loci in a
genome-wide fashion, thus leading to the creation of contact maps. The
democratization of such protocols creates many new opportunities for
methodological development: how to infer 3D models from these contact maps?
Can such models help us gain insights in biological process?
\item Here, I review recent methodological developments \fixme{finish this} 
\end{itemize}

\end{abstract}

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwujekwe:do}.
Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host.

@@

While safe and effictive vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

\section*{Three-dimensional models of DNA from Hi-C data}

@@

Both consensus and ensemble methods have benefits and drawbacks. Ensemble
approaches are more biologically accurate: Hi-C data are derived from a
population of cells, each of those with a uniquely folded 3D structure. The
variability of structures among ours cells may thus be better represented by a
population of structures. Yet, in addition of being more complex and
computationally intensive, ensemble approaches raise the question of
interpretability. Often, one has to fall back to interpreting the mean
structure \citep{kalhor:genome}, or a reduced set of structures
\citep{rousseau:three}. On the other hand, consensus approaches yield a single
structure recapitulating the rich information provided by Hi-C data, including
hallmarks of genome architecture shared across all cells despite the
cell-to-cell variability \citet{nagano:single-cell}. More amenable for
visualization and analysis, this average structure can be easily integrated
with other sources of data, such as RNA-seq, chip-seq etc, which are also
population-based.

@@

In addition of having subtle differences in the derivation of wish distances
(which can have important effects on the resulting structures),
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. Table~\ref{table:mds_detail}
summarizes the characteristic of each method.

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
problem, after filtering interaction counts based on significance and
interpolation of missing values, \citet{ben-elazar:spatial} casts the
inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
the 3D structure is inferred jointly with the wish-distances. Another idea is
to parametrize the count-to-distance mapping as a power-law ($d = \beta
c^\alpha)$, and to infer jointly the parameters jointly with the structure
\citep{zhang:inference}.

@@

\subsubsection*{Statistical-based models}

\citet{varoquaux:statistical} proposes a novel approach called \texttt{Pastis}
based on statistical modeling of contact counts as random Poisson variables
where the intensity is a function of the distance: $c_{ij} \sim
\text{Poisson}(\beta d_{ij}^\alpha)$, and some perform the optimization only
including a subset of contact counts. The authors propose to cast the
inference of 3D models as maximizing the likelihood:

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio, through a direct modeling of the
noise through the statistical model (See figure~\ref{fig:3D_structures}.

\begin{figure}
\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
\label{fig:3D_structures}
\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
(\textbf{C} and \textbf{D}) on two
HiC data set of {\em P. falciparum}. Each color corresponds to a chromosome.
Large white beads mark centromeres, small blue beads telomeres and green beads
VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
cluster is marked with a dashed line.}
\end{figure}

@@

\subsubsection*{Estimating the posterior distribution}

Similarly to \citet{varoquaux:statistical}, \citet{rousseau:three} and
\citet{hu:bayesian} proposes to model contact counts with a formal
probabilistic model. \citet{rousseau:three} models observed contact counts
$c_{ij}$ as a random Gaussian variable of mean $\beta d_{ij}^{\alpha}$,
$\alpha \leq 0$ and variance $\sigma_{ij}$ estimated directly from the contact
count data, while \citet{hu:bayesian} models contact counts as random Poisson
variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
posterior using MCMC. Obtaining a consensus structure from such a method can
be accomplished by selecting the maximum a posteriori.

@@

The last category of data-driven methods to infer the 3D architecture of the
genome rely on a new protocol to probe single-cell for their 3D structures
\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its
early days, and, despite potential for assessing the variability of
cell-to-cell genome architecture in a genome-wide fashion, only a handful of
data sets are today publicly available. The contact maps originating from
these data sets are very sparse, and specific methods to infer 3D structures
need to be developed specifically for sc-HiC. Two methods have been developed
to tackle the challenging task of constructing models from this data.

Akin to {\em consensus} MDS-like method and {\em ensemble}-local minima
methods,  the first approach is to consider each contact as a constraint and
to formulate an under constrained optimization problem. A population of
structures satistifying the constraints can be found by sampling local
minimas.

@@

@@

\subsection*{Model evaluation}

A substantial difficulty in modeling the 3D structure of the genome is that
model evaluation tends to the subjective. What is the relevant measure?
``Truth'' is generally not fully available, except for a few pairs of loci or
in simulations. Is validating the colocalization of a pair or a few pairs of
loci via FISH experiments enough? Is fit to the contact maps or agreements
between modeling techniques relevant?

\section*{Model-driven approach: modeling DNA as a flexible fibers under constraints}

``data-driven'' methods, as presented in the last section, use the
experimental contact maps to infer models as consistent as possible with the
data. Another approach consists in modeling chromosomes as polymers under a
small number constraints such that a generated contact maps from these models
match as closely as possible the observed contact maps. These ``model-driven''
approaches offer mechanistical insights into the genome architecture.

\subsection*{A small set of constraints explain the 3D architecture of \textit{S. cerevisiae}}

@@

\subsection*{Volume-exclusion modeling cannot explain {\em P. falciparum}'s
genome architecture}

For at least some stages, {\em P. falciparum} share a lot of common genomic
architecture hallmarks with the budding yeast: the centromeres are strongly
co-localized at one end of the nucleus, telomeres are in physical contacts,
\dots So could the primary architectural features also arise from a population
of constrained but otherwise random of structures? Adapting the set of
constraints to match biological knowledge of {\em P. falciparum},
\citet{ay:three-dimensional} showed the resulting simulated VE map not only
yielded lower correlations, but also did not show the same features as the
original contact count matrix. In particular, the VRSM genes which display
domain-like enrichment in interactions did not appear in the simulated VE map
(See Figure~\ref{fig:ve_modeling})

\begin{figure}
\caption{Volume Exclusion Modeling}{}
\label{fig_ve_modeling}
\end{figure}

@@

In previous section, I have reviewed data-driven methods to infer either
\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
may well ask why go through the effort to obtain such model and not directly
study the contact maps?  \fixme{why?}In this section, I will review a number of downstream
analysis one can do on 3D models, highlighting but not limiting myself to results
on the 3D structure of {\em P. falciparum}. \fixme{why}

>>>>

\begin{abstract}
The development of new means to probe samples for the three-dimensional
structure of DNA paves the way for in depth and systematic analysis of genome
architecture, offering new insights in many biological processes. 3C-like
methods can now assess physical interactions between pairs of loci in a
genome-wide fashion, thus leading to the creation of contact maps. The
democratization of such protocols creates many new opportunities for
methodological development: how to infer 3D models from these contact maps?
Can such models help us gain insights in biological process?

Recent studies assessed {\em P. falciparum} (the deadliest of the 5
human malaria parasites) for its genome organization using 3C-like protocol.
With its small genomic size, fairly simple yet changing genomic organization
across its lifecyle and strong correlation between chromatin folding and gene
expression, this parasite is the ideal case study for applying and developing
methods to infer 3D models and use them for downstream analysis.

Here, I review a set of methods used to build and analyse three-dimensional
models from contact maps data with a special highlight findings on {\em P.
falciparum}'s genome organization.
\end{abstract}

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwujekwe:do}. The disease is caused by small
unicellular protozoan parasite {\em Plasmodium} that infects its host through
a mosquito bite. Out of the 5 species that can infect human, {\em P.
falciparum} is by the far the most deadly with 99\% of the deaths associated
to it. The parasite has a complex life cycle, with multiple stages both in the
human and mosquito host.

@@

While effective vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at
the epigenetic and post-translational levels \citep{cui:chromatin-mediated,
duffy:role, hoeijmakers:placing, horrocks:control, deitsch:mechanisms}:
a better understanding of genome architecture of {\em P. falciparum}, both at
a local and global scale may provide new leads for developing new therapies.

@@

\section*{``Data-driven'' three-dimensional models of DNA from contact maps}

@@

Both consensus and ensemble methods have benefits and drawbacks. Ensemble
approaches are more biologically accurate: Hi-C data are derived from a
population of cells, each of those with a uniquely folded 3D structure. The
variability of structures among ours cells may thus be better represented by a
population of structures. Yet, in addition of being more complex and
computationally intensive, ensemble approaches raise the question of
interpretability. Often, one has to fall back to interpreting the mean
structure \citep{kalhor:genome}, or a reduced set of structures
\citep{rousseau:three}. On the other hand, consensus approaches yield a single
structure recapitulating the rich information provided by Hi-C data, including
hallmarks of genome architecture shared across all cells despite the
cell-to-cell variability \citep{nagano:single-cell}. More amenable for
visualization and analysis, this average structure can be easily integrated
with other sources of data, such as RNA-seq, chip-seq etc, which are also
population-based. On the other hand, interpretation of \textit{consensus}
model need to be handle with care: these structures should not be considered
as a true representation of the chromatin folding in the nucleus, but as a
summary model of the contact maps.

@@

In addition of having subtle differences in the derivation of wish distances
(which can have important effects on the resulting structures),
the different methods also vary by the inclusion of weights to reflect
confidence in ``wish-distances'' \citep{ay:three-dimensional}, or constraints
to reflect prior knowledge on the structure. See table~\ref{table:mds_detail}
for a summary of the characteristic of each method.

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
problem, after filtering interaction counts based on significance and
interpolation of missing values, \citet{ben-elazar:spatial} casts the
inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
the 3D structure is inferred jointly with the wish-distances. Another idea is
to parametrize the count-to-distance mapping as a power-law ($d = \beta
c^\alpha)$, and to infer jointly the parameters jointly with the structure
\citep{zhang:inference, varoquaux:statistical}.

@@

\subsubsection*{Statistical models for contact counts}

\begin{figure}
\includegraphics[width=\linewidth]{schemas/3D_structures.pdf}
\caption{3D models of {\em P. falciparum}'s genome architecture}{Using
\texttt{pastis-MDS} (\textbf{A} and \textbf{B}) and \texttt{pastis-PO}
(\textbf{C} and \textbf{D}) on two
HiC data set of {\em P. falciparum}: a low signal-to-noise ratio data set
at a {\em Ring} stage \citep{lemieux:genome-wide}
(\textbf{A} and \textbf{C}) and a high signal-to-noise ratio data set
at the {\em Trophozoit} stage \citep{ay:three-dimensional}
(\textbf{B} and \textbf{D}). \texttt{pastis-MDS} recovers a plausible
structure for the high quality data set, but the non-uniform distribution of
beads on low quality data set suggests that the method is not adequate on such
data set. In contrast, \texttt{pastis-PO} recovers plausible structures both
on the high and low quality data sets.
Each color corresponds to a chromosome.
Large white beads mark centromeres, small blue beads telomeres and green beads
VRSSM clusters. Centromeric cluster is marked with a black line, telomeric
cluster is marked with a dashed line.
}
\label{fig:3D_structures}
\end{figure}

Another approach consists of modeling contact counts as random variables, with
the 3D model being a latent variables. For example,
\citet{varoquaux:statistical} proposes an approach called
\texttt{Pastis}, that models contact counts as random Poisson variables where
the intensity of the process between $i$ and $j$ is a function of the
distance: $c_{ij} \sim \text{Poisson}(\beta d_{ij}^\alpha)$. The inference can
thus be cast as maximizing the likelihood:

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio, through a direct modeling of the
noise through the statistical model (See figure~\ref{fig:3D_structures}).

@@

\subsubsection*{Estimating the posterior distribution of a statistical model}

Similarly to \citet{varoquaux:statistical}, \citet{rousseau:three} and
\citet{hu:bayesian} proposes to model contact counts with a formal
probabilistic model. \citet{rousseau:three} models observed contact counts
$c_{ij}$ as a random Gaussian variable of mean $\beta d_{ij}^{\alpha}$,
$\alpha \leq 0$ and variance $\sigma_{ij}$ estimated directly from the contact
count data, while \citet{hu:bayesian} models contact counts as random Poisson
variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
posterior using MCMC. Obtaining a consensus structure from such a method can
be accomplished by selecting the maximum {\em a posteriori}.

@@

The last category of data-driven methods to infer the 3D architecture of the
genome rely on a new protocol to probe single-cell for their 3D structures
\citep{nagano:single-cell,ramani:massively}. Single-cell Hi-C is still in its
early days, and, despite potential for assessing the variability of
cell-to-cell genome architecture in a genome-wide fashion, only a handful of
data sets are today publicly available. The contact maps originating from
these data sets are very sparse, and specific methods to infer 3D structures
need to be developed specifically for sc-HiC.

Akin to {\em consensus} MDS-like method and {\em ensemble}-local minima
methods,  the first approach is to consider each contact as a constraint and
to formulate an under constrained optimization problem. A population of
structures satisfying the constraints can be found by sampling local
minimas.

@@

\fixme{hirata et al}

@@

\subsection*{Model evaluation and comparison}

A substantial difficulty in modeling the 3D structure of the genome is that
model evaluation tends to the subjective. What is the relevant measure?
``Truth'' is generally not fully available, except for a few pairs of loci or
in simulations. Is validating the colocalization of a pair or a few pairs of
loci via FISH experiments enough? Is fit to the contact maps or agreements
between modeling techniques relevant? Are the conclusion drawn from 3D models
in agreement when these are inferred from different methods?

First, methods can be validated and compared against contact maps simulated
from a known ground truth \citep{zhang:inference, varoquaux:statistical,
lesne:3d, trieu:3D}. Note that while this is a simple and natural first step
for methods inferring a \textit{consensus} structure, comparing and assessing
robustness and accuracy of {\em ensemble} of 3D models is much more
challenging: in fact, it is still an untackled problem. Second, one can assess
the stability and robustness of the inference with respect to (1) data
bootstrapping; (2) contact map resolution (the models should not change as the
resolution of the data varies) \citep{zhang:inference, varoquaux:statistical};
(3) and in between biological replicates \citep{zhang:inference,
varoquaux:statistical}. As a cautionary remark, it is important to stress that
it is not because a method is very stable that is ``good'' with respect to any
criterion: I can imagine a number of very stable methods, that would yet
provide absolutely no insights on genome organization. Third, models can be
compared to other sources of data, such as FISH
\citep{ay:three-dimensional,duan:three-dimensional}. Fourth, biological
plausibility of the resulting models can be considered: are the beads
uniformly distributed in the cell \citep{ay:three-dimensional}? are known
hallmarks of the genome architecture such as centromeres clustering preserved?

These are a handfull of ways to assess plausibility and accuracy of 3D
reconstruction, but many avenues in model evaluation and comparison are yet to
be explored.

\section*{The art of modeling genome architecture.}

``Data-driven'' methods, as presented in the last section, use the
experimental contact maps to infer models as consistent as possible with the
data. ``Model-driven'' tackle the 3D-modeling challenge exactly in the
opposite manner. Consider the following task. You have a tens of thousands of
randomly placed beads-on-strings. Can you find the smallest sets of
constraints such that these beads-on-a-string interact overall the same way as
a given contact map? This is the daunting task accomplished by
``model-driven'' approach: chromosomes are modeled as polymers (or random
self-excluding fibers) under a small number constraints such that a generated
contact maps from these models match as closely as possible the observed
contact maps. These ``model-driven'' approaches offer mechanistical insights
into the genome architecture.

\subsection*{Building a yeast nucleus}

@@

\subsection*{Building a {\em P. falciparum} nucleus?}

\begin{figure}
\caption{Volume Exclusion Modeling}{}
\label{fig:ve_modeling}
\end{figure}

For at least some stages, {\em P. falciparum} share a lot of common genomic
architecture hallmarks with the budding yeast: the centromeres are strongly
co-localized at one end of the nucleus (for at least some stages),
telomeres are in physical contacts with one another,
\dots So could the primary architectural features also arise from a population
of constrained but otherwise random of structures? Adapting the set of
constraints to match biological knowledge of {\em P. falciparum},
\citet{ay:three-dimensional} showed the resulting simulated VE map not only
yielded lower correlations, but also did not show the same features as the
original contact count matrix. In particular, the VRSM genes which display
domain-like enrichment in interactions do not appear in the simulated VE map
(See Figure~\ref{fig:ve_modeling}). So\dots Can we add constraints on VRSM
gene clusters to improve correlations between true and generated contact maps?
Adding a constraint on all beads considered as VRSM clusters is not enough:
running 100 experiments yielded structures that did not fulfill all the
constraints, demonstrating VRSM genes cannot all cluster together in a cell.

@@

In previous section, I have reviewed data-driven methods to infer either
\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
may well ask why go through the effort to obtain such model and not directly
study the contact maps?  In this section, I will review a number of downstream
analysis one can do on 3D models, highlighting but not limiting myself to results
on the 3D structure of {\em P. falciparum}.
